---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First I will clean any remaining variable in the environment:
```{r}
rm(list = ls(all.names = TRUE))
```
Then if the **activity.csv** file is not present in the current directory, one has to download it from [Activity Dataset for Assignment 1](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) and place it in the current directory.
As described in the exercise, this dataset has recorded **steps** in a 5-minute **interval** for a specified **date**.
I load the file, trasform the date into a date object and do some preliminary inspection:
```{r}
activity = read.csv("activity.csv", header = T, stringsAsFactors = FALSE)
activity$date = as.Date(activity$Date)
str(activity)
head(activity)
summary(activity)
```


## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
